---
title: "Exercise set 5"
author: "Other template for references"
date: "28/11/2021"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Exercise 1. Creating graphs and visualization

```{r E1-Q2.1}
## Add your code here
library(graph)
#library(help="graph")
?graphNEL
#gR <- data("MAPKsig")
#gR

numOfNodes= numNodes(MAPKsig)
numOfEdges= numEdges(MAPKsig)

data(MAPKsig)
adjBAD<-adj(MAPKsig,"BAD")

outDegree <- degree(MAPKsig,"RALA")

numOfNodes
numOfEdges
outDegree
adjBAD

#if (require(Rgraphviz)) {
 #  nat = rep(FALSE, length(nodes(MAPKsig)))
  # names(nat) = nodes(MAPKsig)
   #plot(MAPKsig, nodeAttrs=list(fixedsize=nat))
 #  }
#edgeWeights(gR)
#signature(object = "graphNEL")
#numOfNodes= numNodes(object = data(MAPKsig))
#numOfEdges= numEdges(data(MAPKsig))
#adj(gR)
#signature(object = data(MAPKsig),adj="BAD")
```



```{r hint-1}
## Creating a graphNEL that shows the distances among 3 major cities in Finland
V <- c("Tampere", "Helsinki", "Turku")
edL = list(Tampere=list(edges=c("Helsinki", "Turku"), weights=c(178, 164)), 
           Turku=list(edges=c("Tampere", "Helsinki"), weights=c(164, 169)),
           Helsinki=list(edges=c("Tampere", "Turku"), weights=c(178, 169))
           )
cities = graphNEL(nodes=V, edgeL=edL)
cities
plot(cities)
```



```{r hint-2}
## If value is equal to 7, we win, everything else we lose!
values = c(1:10)
result = list()
for (value in values){
  if (value == 7) {
    result[[value]] = "Win!"
  }
  else {
    result[[value]] = "Loss!"
  }
}
result
```



```{r E1-Q3.1}
## Add your code here
selected_key="GO:0019395"
selected_key_offspring  <- as.list(GOBPCHILDREN[selected_key])
atest <- as.vector(unlist(selected_key_offspring))
V <- append(atest,"GO:0019395")
E=list()
E$`GO:0019395`$edges = c( "GO:0001561", 
          "GO:0006635",
          "GO:0010430",
          "GO:0046320",
          "GO:0046321",
          "GO:0046322"
          )
# E = list("GO:0001561"=list(edges=c(character(0))), 
#             "GO:0006635"=list(edges=c(character(0))),
#             "GO:0010430"=list(edges=c(character(0))),
#             "GO:0046320"=list(edges=c(character(0))), 
#             "GO:0046321"=list(edges=c(character(0))),
#             "GO:0046322"=list(edges=c(character(0))),
#             "GO:0019395"=list(edges=V))

VV = c("GO:0001561", "GO:0006635","GO:0019395")
EE = list("GO:0001561"=list(edges=c()), 
          "GO:0006635"=list(edges=c()),
          "GO:0010430"=list(edges=c()),
          "GO:0046320"=list(edges=c()),
          "GO:0046321"=list(edges=c()),
          "GO:0046322"=list(edges=c()),
          "GO:0019395"=list(edges=E$`GO:0019395`$edges)
          
    
          
          )
            
# 
# length(V)
# E
# E<-unlist(E)
# length(E)

G = graphNEL(nodes=V, edgeL=EE, edgemode="directed")
G
degree(G)
plot(G)

```
```
#V <- select(GO.db, keys = selected_key_offspring[[1]][1:10], keytype = "GOID", columns = "TERM")
# selected_key_offspring <- unlist(selected_key_offspring[1:6])
# row.names(selected_key_offspring)<-NULL
# 
# selected_key_offspring[7] = selected_key
# selected_key_offspring
#V<-V$GOID
#V
#E = list()
#for (value in V){
#  if (V[value] == "GO:0019395") {
#  E = list("GO:0019395"=list(edges=E$GO:0019395$edges))
#       }
#  else {
#  E = list("GO:0019395"=list(edges=character(0)))
#  }
#}

#G = graphNEL(nodes=as.character(V), edgeL=E)
#G
#plot(G)


```



## Exercise 2. Pathway enrichment analysis (MAPK signaling pathway)

```{r E2-Q1.1}
## Add your code here
count_matrix <- readRDS("/student_data/BBT.BI.202_2021/data/week_5/count_matrix_normalized_logged.RDS")
gene_list <-  readRDS("/student_data/BBT.BI.202_2021/data/week_5/DE_genes_PC_vs_BPH.RDS")

```



```{r E2-Q1.2}
## Add your code here
ensembl_gene_ids <- row.names(count_matrix)
to_convert = ensembl_gene_ids 
to_convert <- na.omit(to_convert)
library('biomaRt')
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

mappings_ensembl_entrez <- getBM(
  filters= "ensembl_gene_id", 
  attributes = c("ensembl_gene_id","entrezgene_id"),
  values=to_convert, 
  mart= mart) 


```



```{r E2-Q1.3}
library(KEGG.db)
#a<-class(KEGGEXTID2PATHID)
#a
#KEGGEXTID2PATHID
map_ke <- as.list(KEGGEXTID2PATHID)
#map_ke
#!is.na(match(mappings_ensembl_entrez$entrezgene_id, names(map_ke)))
kegg_entrez_pathway = map_ke[!is.na(match(mappings_ensembl_entrez$entrezgene_id, names(map_ke)))]
#typeof(kegg_entrez_pathway)
ensembl2kegg_pathway = list()
#a <- class(mappings_ensembl_entrez[, ])
#a
#ss<- typeof(mappings_ensembl_entrez[i, ]$entrezgene_id)
#ss
#typeof(temp)
#sss<-nrow(mappings_ensembl_entrez)
for (i in 1:nrow(mappings_ensembl_entrez)) {
  temp = kegg_entrez_pathway[as.character(mappings_ensembl_entrez[i, ]$entrezgene_id)]
  if (!is.na(names(temp))){
    ensembl2kegg_pathway[mappings_ensembl_entrez[i, ]$ensembl_gene_id] = temp
  } 
}

for (i in 1:nrow(mappings_ensembl_entrez)) {
  temp = kegg_entrez_pathway[as.character(mappings_ensembl_entrez[i, ]$entrezgene_id)]
  if (!is.na(names(temp))){
    ensembl2kegg_pathway[mappings_ensembl_entrez[i, ]$ensembl_gene_id] = temp
  } 
}
#test
#atestInheritedMethods()a <- typeof(temp)
#a
```



### Pathway enrichment

```{r E2-Q1.4}
## Add your code here
#library(help="KEGG.db")
#KEGGEXTID2PATHID
#library(KEGG.db)
KEGGPathways <- as.list(KEGGPATHNAME2ID)

#descriptions <- sapply(xx, function(x) unlist(attributes(get(x))))

#attributes(xx)

KEGGPathwaysNames <- as.matrix(unlist(list(attributes(KEGGPathways))))
#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
KEGGPathways[grep("MAPK signaling pathway",KEGGPathwaysNames)]
#toString(grep("MAPK signaling pathway",attributes(xx)))
#select(,)         
```



#### Calculate N

```{r hint-3}
l = list(g1=c('p1'),
         g2=c('p2', 'p3'), # g2 is assigned to two ps.
         g3=c('p1'),
         g4= NA   # g4 is not assigned to any p.
         )
l
unlist(l) # check how the gene names are modified
sum(!is.na(unlist(l)))
```




```{r E2-Q1.6}
## Add your code here
#KEGGPathways
#unlist(ensembl2kegg_pathway) # check how the gene names are modified
N <- sum(!is.na(unlist(ensembl2kegg_pathway)))
N
```



#### Calculate n for our pathway

```{r hint-4}
l = list(g1=c('p1'),
         g2=c('p2', 'p3'), # g2 is assigned to two ps.
         g3=c('p1'),
         g4= NA   # g4 is not assigned to any p.
         )

l_table = table(unlist(l))
l_table
l_table["p3"]
```



```{r E2-Q1.8}
## Add your code here

l_table = table(unlist(ensembl2kegg_pathway))
#l_table
n <- l_table["hsa04010"]
n
```



#### Calculate M

```{r E2-Q1.10}
## Add your code here

ensembl2kegg_pathway_de <- ensembl2kegg_pathway[names(ensembl2kegg_pathway) %in% rownames(gene_list)]
M <- sum(!is.na(unlist(ensembl2kegg_pathway_de)))
M
```



#### Calculate k for our pathway

```{r E2-Q1.12}
## Add your code here
l_table = table(unlist(ensembl2kegg_pathway_de))
#l_table
k <- l_table["hsa04010"]
k
```



#### Calculate enrichment P-values

```{r E2-Q1.14}
## Add your code here
#sum(phyper(k-1, M, N-M, n))
sum(dhyper(k:n, M, N-M, n))
1- phyper(k-1, M, N-M, n)
N
M
```



## Exercise 3. Pathway visualization

#### Calculate mean difference between PC and BPH used by pathview i.e. `gene.data`

```{r E3-Q2.1}
## Add your code here
#library(help="pathview")

meanBPH <- apply(count_matrix[,1:5], 1, sum)/5 
meanPC <- apply(count_matrix[,6:20], 1, sum)/15
count_matrix_mean_diff <- (meanPC-meanBPH)
which.min(count_matrix_mean_diff)
which.max(count_matrix_mean_diff)
min(count_matrix_mean_diff)
```



```{r E3-Q2.4}
## Add your code here
library(pathview)

#a <- grep("hsa04010",ensembl2kegg_pathway_de)
#genes_MAPK_signaling_pathway <- unlist(attributes(ensembl2kegg_pathway_de[a]))

cc<-as.matrix(count_matrix_mean_diff)
pathviewOutput <- pathview(gene.data=cc, pathway.id="04010", species="hsa", gene.idtype="ENSEMBL",out.suffix = "MAPK_signaling", limit = list(gene=2, cpd=1),kegg.native = TRUE)
```



## Exercise 4. R script interpretation




## Exercise 5. Pathway enrichment analysis; many pathways

### Calculate n for all pathways

```{r E5-Q1.1}
## Add your code here
N
M
ns <- table(unlist(ensembl2kegg_pathway))
#l_table
#n <- l_table["hsa04010"]
#n
#KEGGPathwaysNames
#for (i in KEGGPathwaysNames){
# ns[i] = l_table[]
#  }
 ns 
```



### Calculate k for all pathways

```{r E5-Q1.2}
## Add your code here
ks <- table(unlist(ensembl2kegg_pathway_de))
ks
```



### Calculate pathway enrichment P-values

```{r E5-Q1.3}
## Add your code here
ps=matrix()
for(i in 1:length(ks)){
#ps[i]= 1- phyper(ks[i]-1, M, N-M, ns[i])
ps[i]=sum(dhyper(ks[i]:n, M, N-M, ns[i]))

}
#ps
min(ps)

```


### Significant pathways

```{r E5-Q1.5}
## Add your code here
#rownames(ks)
ps2<- as.matrix(ps)
row.names(ps2) = rownames(ks)
sum(length(which(ps<0.05)))
#ps2
```



```{r E5-Q1.7}
## Add your code here
#rownames(ps2)
a=ps2[(which(ps2[,1]<0.05)),]
a<-sort(a,decreasing = FALSE)
a<-data.frame(a)
a
# a<-a[order(a$a),]
# a<-data.frame(a)
# a
```



```{r E5-Q1.9}
## Add your code here
```
