---
title: "Exercise set 5"
author: "Hien Le"
date: " 25/11/2021"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/student_data/BBT.BI.202_2021/students/leh/exercise_5")

```



## Exercise 1. Creating graphs and visualization

```{r E1-Q2.1}
## Add your code here
library(graph)
library(help="graph")
?graphNEL
data("MAPKsig")
length(nodes(MAPKsig)) #how many nodes here
length(unlist(edges(MAPKsig))) # how many edges here
degree(MAPKsig) # checking out and in degree
```



```{r hint-1}
## Creating a graphNEL that shows the distances among 3 major cities in Finland
V <- c("Tampere", "Helsinki", "Turku")
edL = list(Tampere=list(edges=c("Helsinki", "Turku"), weights=c(178, 164)), 
           Turku=list(edges=c("Tampere", "Helsinki"), weights=c(164, 169)),
           Helsinki=list(edges=c("Tampere", "Turku"), weights=c(178, 169))
           )
cities = graphNEL(nodes=V, edgeL=edL)
cities
plot(cities)
```



```{r hint-2}
## If value is equal to 7, we win, everything else we lose!
values = c(1:10)
result = list()
for (value in values){
  if (value == 7) {
    result[[value]] = "Win!"
  }
  else {
    result[[value]] = "Loss!"
  }
}
result
```



```{r E1-Q3.1}
## Add your code here
library(GO.db)
ll = as.list(GOBPCHILDREN["GO:0019395"]) # all direct children
V = c("GO:0019395", ll$`GO:0019395`) # Include the term itself of our interested GO
V = unname(c("GO:0019395", GOBPCHILDREN$"GO:0019395")) 
E=list()

for (id in V) { E[[id]]$edges = character(0) }
E[[V[1]]]$edges = V[2:length(V)]
names(E) = V
E
names(E) = V
length(V)
G = graphNEL(V, E, edgemode="directed") # "directed" in here will show the trend of interaction
G
plot(G)
degree(G)
```



## Exercise 2. Pathway enrichment analysis (MAPK signaling pathway)

```{r E2-Q1.1}
## Add your code here
count_matrix = readRDS("count_matrix_normalized_logged.RDS")
gene_list = readRDS("DE_genes_PC_vs_BPH.RDS")

```



```{r E2-Q1.2}
## Add your code here
library(biomaRt)
ensembl_id_count_matrix = rownames(count_matrix)
#mapping
to_convert = ensembl_id_count_matrix
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
mappings_ensembl_entrez <- getBM(
                  filters= "ensembl_gene_id", 
                  attributes = c("ensembl_gene_id", "entrezgene_id"),
                  values=to_convert, 
                  mart= mart) 
dim(mappings_ensembl_entrez)
```



```{r E2-Q1.3}
library(KEGG.db)
map_ke <- as.list(KEGGEXTID2PATHID)
kegg_entrez_pathway = map_ke[!is.na(match(mappings_ensembl_entrez$entrezgene_id, names(map_ke)))]

ensembl2kegg_pathway = list()

for (i in 1:nrow(mappings_ensembl_entrez)) {
  temp = kegg_entrez_pathway[as.character(mappings_ensembl_entrez[i, ]$entrezgene_id)]
  if (!is.na(names(temp))){
    ensembl2kegg_pathway[mappings_ensembl_entrez[i, ]$ensembl_gene_id] = temp
  } 
}

#Another way of ensembl12kegg_pathway
library(KEGG.db)
map_ke <- as.list(KEGGEXTID2PATHID)
kegg_entrez_pathway = map_ke[!is.na(match(mappings_ensembl_entrez$entrezgene_id, names(map_ke)))]

index_arr = which(as.character(mappings_ensembl_entrez$entrezgene_id) %in% names(kegg_entrez_pathway))
ensembl2kegg_pathway = list()
ensembl2kegg_pathway[mappings_ensembl_entrez[index_arr, ]$ensembl_gene_id] = kegg_entrez_pathway[as.character(mappings_ensembl_entrez[index_arr, ]$entrezgene_id)]

```



### Pathway enrichment

```{r E2-Q1.4}
## Add your code here
library(help="KEGG.db")
library(KEGG.db)
### Annotation data contains the information for mappings between KEGG pathway names to identifiers : KEGGPATHNAME2ID
object_annotation = as.list(KEGGPATHNAME2ID)
MAPK_idx_list <- grep("MAPK", names(object_annotation))
MAPK_idx_list[1]#261
```



#### Calculate N

```{r hint-3}
l = list(g1=c('p1'),
         g2=c('p2', 'p3'), # g2 is assigned to two ps.
         g3=c('p1'),
         g4= NA   # g4 is not assigned to any p.
         )
l
unlist(l) # check how the gene names are modified
sum(!is.na(unlist(l)))
```




```{r E2-Q1.6}
## Add your code here
ensembl2kegg_pathway
unlist(ensembl2kegg_pathway) # check how the gene names are modified
N = sum(!is.na(unlist(ensembl2kegg_pathway)))
N
```



#### Calculate n for our pathway

```{r hint-4}
l = list(g1=c('p1'),
         g2=c('p2', 'p3'), # g2 is assigned to two ps.
         g3=c('p1'),
         g4= NA   # g4 is not assigned to any p.
         )

l_table = table(unlist(l))
l_table
l_table["p3"]
```



```{r E2-Q1.8}
## Add your code here
ensembl2kegg_pathway_table = table(unlist(ensembl2kegg_pathway))
ensembl2kegg_pathway_table
n_test = ensembl2kegg_pathway_table["hsa04010"]
n = n_test[[1]]
n

```



#### Calculate M

```{r E2-Q1.10}
## Add your code here

names(ensembl2kegg_pathway)
rownames(gene_list)
names(ensembl2kegg_pathway) %in% rownames(gene_list)
ensembl2kegg_pathway_de = ensembl2kegg_pathway[names(ensembl2kegg_pathway) %in% rownames(gene_list)]
unlist(ensembl2kegg_pathway_de) # check how the gene names are modified
M = sum(!is.na(unlist(ensembl2kegg_pathway_de)))
M
```



#### Calculate k for our pathway

```{r E2-Q1.12}
## Add your code here

ensembl2kegg_pathway_de_table = table(unlist(ensembl2kegg_pathway_de))
ensembl2kegg_pathway_de_table
k_test = ensembl2kegg_pathway_de_table["hsa04010"]
k = k_test[[1]]
```



#### Calculate enrichment P-values

```{r E2-Q1.14}
## Add your code here
N
M
n
k
sum(dhyper(k:n, M, N-M, n))
```



## Exercise 3. Pathway visualization

#### Calculate mean difference between PC and BPH used by pathview i.e. `gene.data`

```{r E3-Q2.1}
## Add your code here
library(pathview)
?pathway
BPH_samples = count_matrix[,1:5]
PC_samples = count_matrix[,6:20]
count_matrix_mean_diff = apply(PC_samples,1,mean) - apply(BPH_samples,1,mean) 
which.max(count_matrix_mean_diff)
round(min(count_matrix_mean_diff),2)
```



```{r E3-Q2.4}
## Add your code here
library(pathview)
count_matrix_mean_diff_matrix = as.matrix(count_matrix_mean_diff)
pv.out = pathview(gene.data = count_matrix_mean_diff_matrix,
                  gene.idtype = "ENSEMBL",
                  pathway.id = "04010",
                  out.suffix="MAPK_signaling", 
                  limit=list(gene=2, cpd=1),
                  species = "hsa",
                  kegg.native = TRUE,
                  )

```



## Exercise 4. R script interpretation




## Exercise 5. Pathway enrichment analysis; many pathways

### Calculate n for all pathways

```{r E5-Q1.1}
## Add your code here
ensembl2kegg_pathway_table = table(unlist(ensembl2kegg_pathway))
ensembl2kegg_pathway_de_table = table(unlist(ensembl2kegg_pathway_de))
ns = c()
ks = c()
#ns
ensembl2kegg_pathway_table_filter = ensembl2kegg_pathway_table[names(ensembl2kegg_pathway_table) %in% names(ensembl2kegg_pathway_de_table)]
for( i in 1:length(ensembl2kegg_pathway_table_filter)){
  ns[i] = ensembl2kegg_pathway_table_filter[[i]]
}
ns




```



### Calculate k for all pathways

```{r E5-Q1.2}
## Add your code here
#ks
ensembl2kegg_pathway_de_table = table(unlist(ensembl2kegg_pathway_de))
ensembl2kegg_pathway_de_table
for( i in 1:length(ensembl2kegg_pathway_de_table)){
  ks[i] = ensembl2kegg_pathway_de_table[[i]]
}
ks
length(ns)
length(ks)
```



### Calculate pathway enrichment P-values

```{r E5-Q1.3}
## Add your code here
#pValue
pValues <- c()
for(i in 1:length(ks)){
  pValues[i] = sum(dhyper(ks[i]:ns[i], M, N-M, ns[i]))
}
pValues
min(pValues)
 

```


### Significant pathways

```{r E5-Q1.5}
## Add your code here
names(pValues) = names(ensembl2kegg_pathway_de_table)
sum(pValues < 0.05)
```



```{r E5-Q1.7}
## Add your code here
#df_pValues_sort
df_pValues = data.frame(pathway_identifier = names(ensembl2kegg_pathway_de_table), pValue = pValues)
df_pValues_sort = df_pValues[order(df_pValues$pValue),]
```



```{r E5-Q1.9}
## Add your code here
```

